---
title: "Bike Sharing Analysis, Time Series Final Project"
author: "Derek Rogers and Sakava Kiv"
date: "2024-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tswge)
library(PolynomF)
library(tidyverse)
library(GGally)
library(astsa)
library(vars)
library(nnfor)
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
```

```{r}
# Set the working directory to the specified path
setwd("C:/SMU MS Data Science/TimeSeries/FinalProject/BikeSharingTimeSeriesAnalysis")

# Check to confirm that the working directory has been set correctly
getwd()
```

This dataset can be used to forecast demand to avoid oversupply and shortages. It spans from January 1, 2011, until December 31, 2018. Determining new station locations, analyzing movement patterns or planning routes will only be possible with additional data.

date - date with the format yyyy-mm-dd
temp_avg - average daily temperature in degree Celsius
temp_min - minimum daily temperature in degree Celsius
temp_max - maximum daily temperature in degree Celsius
temp_observ - temperature at the time of observation in degree Celsius
precip - amount of precipitation in mm
wind - wind speed in meters per second
wt_fog - weather type fog, ice fog, or freezing fog (may include heavy fog)
wt_heavy_fog - weather type heavy fog or heaving freezing fog (not always distinguished from fog)
wt_thunder - weather type thunder
wt_sleet - weather type ice pellets, sleet, snow pellets, or small hail
wt_hail - weather type hail (may include small hail)
wt_glaze - weather type glaze or rime
wt_haze - weather type smoke or haze
wt_drift_snow - weather type blowing or drifting snow
wt_high_wind - weather type high or damaging winds
wt_mist - weather type mist
wt_drizzle - weather type drizzle
wt_rain - weather type rain (may include freezing rain, drizzle, and freezing drizzle)
wt_freeze_rain - weather type freezing rain
wt_snow - weather type snow, snow pellets, snow grains, or ice crystals
wt_ground_fog - weather type ground fog
wt_ice_fog - weather type ice fog or freezing fog
wt_freeze_drizzle - weather type freezing drizzle
wt_unknown - weather type unknown source of precipitation
casual - number of unregistered customers
registered - number of registered customers
total_cust - sum of registered and casual customers
holiday - indicates whether the day is a holiday or not

Acknowledgements
data used to create this dataset was taken from:

Capital Bikeshare for the bike sharing demand,
NOAA's National Climatic Data Center for weather data,
DC Department of Human Resources for data on public holidays.

The chosen data set is the bike_sharing_dataset.csv from kaggle: https://www.kaggle.com/datasets/juliajemals/bike-sharing-washington-dc

The chosen response (target) variable is total_cust

Being able to model and predict total_cust can help entities directly involved or adjacent to bike sharing better understand and predict the varying demand to avoid oversupply or shortages. Forecasting future demand can help plan bike maintenance as heavily used bikes break down more often. This information can help inform business decisions, setting prices, and creating effective advertisements. 

```{r}
# The chosen data set is the bike_sharing_dataset.csv from kaggle: https://www.kaggle.com/datasets/juliajemals/bike-sharing-washington-dc
# The chosen response variable is total_cust
# Being able to model and predict total_cust can help entities directly involved or adjacent to bike sharing better understand and predict the varying demand to avoid oversupply or shortages. Forecasting future demand can help plan bike maintance as heavily used bikes break down more often. This information can help inform business decisions, setting prices, and creating efective advertisements. 

BikeSharing = read.csv("bike_sharing_dataset.csv", header = TRUE)
BikeSharing_train = data.frame(TotalCust = ts(BikeSharing$total_cust))

head(BikeSharing)
nrow(BikeSharing)

# There are 4 consecutive missing values
which(is.na(BikeSharing$total_cust))
```

```{r}
# Impute the 4 consecutive missing values with the average of the value before and after the missing values

# Calculate the average of the values at positions 1848 and 1853
average_value = mean(c(BikeSharing$total_cust[1848], BikeSharing$total_cust[1853]), na.rm = TRUE)
average_value

# Assign the average value to the missing positions 1849, 1850, 1851, and 1852
BikeSharing$total_cust[1849:1852] = average_value

# Optionally, you can verify the imputation by checking the values at these positions
BikeSharing$total_cust[1848:1853]

# Plot around where the imputed values are
plot(BikeSharing$total_cust[1838:1863], type = "l")
```

```{r}
# Plot the full dataset

# The realization plot shows strong periodic trends, wandering behavior, and variance that changes over time.

# We see slowly dampening autocorrelation with weak and an irregular length periodic behavior.

# There is a peak at zero in the parzen frequency window.
# The parzen window also shows a peak around 0.15, which is a period of 1/0.15 = 6.666. 
#   B/c this data is collected daily, there evidence to suggest a weekly trend.
#   We can hypothesize total_cust could have a greater total_cust on the weekends when all other variables are held constant. 
plotts.sample.wge(BikeSharing$total_cust, arlimits = TRUE)
```

```{r}
# The CCF plot shows a log of 1 for precipitation. This makes sense a we might expect less people to rent a bike the day after it rains.
ccf(BikeSharing$precip, BikeSharing$total_cust)
```

```{r}
BikeSharing$precip_L1 = dplyr::lag(BikeSharing$precip, 1)
BikeSharing$precip_L1[is.na(BikeSharing$precip_L1)] = median(BikeSharing$precip_L1, na.rm = TRUE)
```

```{r}
# Shows the present and missing values are represented consistently throughout the plot, giving evidence the NA value can be translated to 0 and be interpreted as non holidays.
data <- data.frame(x = is.na(BikeSharing$holiday))
gg = ggplot(data, aes(x = 1:nrow(data), y = x)) +
  geom_point(size = 1, height = .5, width = .5, color = "red") +
  xlab("days by row number") +
 ylab("wt_rain Values Present or Missing") +
 labs(title = "Scatter Plot of Missing wt_rain Values", size = 10) +
 scale_y_discrete(labels = c("Present", "Missing")) +
 scale_x_continuous(breaks = seq(0, nrow(data), by = 100)) +
 theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
 theme_classic()
gg
```

```{r}
# Replace NA with 0 which represents non holidays, 1 represents holidays
BikeSharing$holiday = replace_na(BikeSharing$holiday, 0)
head(BikeSharing$holiday)
```

```{r}
ggpairs(BikeSharing[, c(28,3,4,5,6,7)])
```

We will try to experiment to estimate for candidate AR or ARMA model based on AIC and BIC values here. We can observe that AN AR(8) model was picked by both AIC AND BIC and also by the PACF function (Box and jenkins method) we saw evidence of an AR(8) as well.

```{r echo = FALSE}
# Top AIC was p = 9, q = 4
aic5.wge(BikeSharing$total_cust, p = 0:12, q = 0:4)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:0)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:0)


# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:12, q = 0:4,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:2,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:0,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:2,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:2, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:0,type = "bic")

pacf(BikeSharing$total_cust) # suggest an AR(8) model
```

```{r}
# transform data
# Difference the data to until we get a stationary model to check for (1-B)^d
# we could use the Dickey-Fuller Test he to see if model as evidence for a unit root, but differencing here I do not see visual evidence for a (1-B) term.
b1=artrans.wge(BikeSharing$total_cust,phi.tr=1)
#although the data does not look white we do see in the acf that a sinusoidal trend appears
b7=artrans.wge(BikeSharing$total_cust,phi.tr = c(rep(0,6),1))

# From the below plots there does not seem to be evidence for a (1-B)^d So we will not include this in the model. 
# The Autocorrelation Function (ACF) plot does not suggest white noise. In a white noise series, we would expect most of the spikes in the ACF plot to fall within the confidence interval (usually represented by blue shaded areas or dashed lines), indicating that there is no significant correlation at different lags.

# However, the acf indicates that there are significant correlations at various lags. This suggests that the time series data exhibits some patterns or relationships between its values at different times, which is not characteristic of white noise.

plotts.sample.wge(b1, arlimits = TRUE)
plotts.sample.wge(b7, arlimits = TRUE)
```

```{r}
# Transform data based off of Sakavas analysis
b7 = artrans.wge(BikeSharing$total_cust, phi.tr = c(rep(0, 6), 1))
b7
b365 = artrans.wge(BikeSharing$total_cust, phi.tr = c(rep(0, 364), 1))
b365
b365_7 = artrans.wge(b365, phi.tr = c(rep(0, 6), 1))
b365_7

plotts.sample.wge(b365_7, arlimits = TRUE)
```

Here we will choose an AR(8) still because of the AIC and BIC balance and Box and Jenkins method. And because AR(8) is a simple model.

```{r echo = FALSE}
# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:12, q = 0)
# Top AIC was p = 7, q = 2
aic5.wge(b7, p = 0:9, q = 0:2)
# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:9, q = 0:1)
# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:9, q = 0:0)
# Top AIC was p = 7, q = 2
aic5.wge(b7, p = 0:8, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(b7, p = 0:8, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:0)

# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:12, q = 0,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:2,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:0,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:2,type = "bic")
# Top BIC was p = 1, q = 0
aic5.wge(b7, p = 0:2, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:0,type = "bic")
```

# ARIMA Modle

```{r}
fit_7 = est.ar.wge(b7, p = 8)

pred_arima_7 = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_7$phi, n.ahead = 7,
    lastn = TRUE, limits = FALSE)

plotts.sample.wge(fit_7$res, arlimits = TRUE)
ljung.wge(fit_7$res, p = 8)
```

```{r}
pred_arima_182 = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_7$phi, n.ahead = 182,
    lastn = TRUE, limits = FALSE)
```

```{r}
ASE_7 = mean((tail(BikeSharing$total_cust, 7) - pred_arima_7$f)^2)
ASE_7

ASE_182 = mean((tail(BikeSharing$total_cust, 182) - pred_arima_182$f)^2)
ASE_182
```

```{r}
plotts.sample.wge(fit_7$res, arlimits = TRUE)
```

```{r}
roll.win.rmse.wge(b7, phi = fit_7$phi, horizon = 7)
roll.win.rmse.wge(b7, phi = fit_7$phi, horizon = 182)
```

```{r}

#Forecasting with ARUMA model here such that ARUMA(p,d,q) with s=7
f = fore.arima.wge(BikeSharing$total_cust, s=7, phi = fit_7$phi, n.ahead = 182, limits =  F, lastn = T, plot=TRUE)

# Plot the actual vs predicted 'total_cust'
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted 26 weeks", xlab = "Time", ylab = "Total Customers", xlim = c(2500, nrow(BikeSharing)))
lines((nrow(BikeSharing) - 181):nrow(BikeSharing), f$f, col = "red")

#Forecasting with SARIMA model here such that ARIMA(p,d,q) with s=7
f2 = fore.arima.wge(BikeSharing$total_cust, s=7, phi = fit_7$phi, n.ahead = 7, limits =  F, lastn = T, plot=TRUE)

# Plot the actual vs predicted 'total_cust'
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted one week", xlab = "Time", ylab = "Total Customers", xlim = c(2800, nrow(BikeSharing)))
lines((nrow(BikeSharing) - 6):nrow(BikeSharing), f2$f, col = "red")
```

### Additional methods to check to see if (1-B^7) term makes sense and is useful
- Overfitting method
- overlaying plots while comparing for Spectral density, ACF, and realizations

```{r}

#compare overfit estimated table with the season factor table of S =7 => observe that when comparing the tables we see some evidence of an S=7, although the estimates are not that strongly matching the (1-B^7) factor tables
esttable=est.ar.wge(BikeSharing$total_cust,p = 12,method='burg') #overfit for seasonality s= 7
# estimate parameters of stationary part
b7table = factor.wge(c(rep(0,6),1)) #1-B^7 table
#factor.wge(c(rep(0,9),1)) #1-B^10


###################################

#Compare Spectral Densities
sims = 30
SpecDen = parzen.wge(BikeSharing$total_cust, plot = "FALSE")
plot(SpecDen$freq,SpecDen$pzgram, type = "l", lwd = 6)

for( i in 1: sims)
{
   SpecDen2 = parzen.wge(gen.arima.wge(319, s=7, phi = fit_7$phi, plot ="FALSE"), plot ="FALSE")
   lines(SpecDen2$freq,SpecDen2$pzgram, lwd = 2, col = "red")
}


#Compare ACFs
sims = 30
ACF = acf(BikeSharing$total_cust, plot = "FALSE")
plot(ACF$lag ,ACF$acf , type = "l", lwd = 6)

for( i in 1: sims)
{
   ACF2 = acf(gen.arima.wge(319, s = 7, phi = fit_7$phi, plot ="FALSE"), plot ="FALSE")
   lines(ACF2$lag ,ACF2$acf, lwd = 2, col = "red")
}

#Compare Generated Realizations 
S10AR3gen = gen.arima.wge(319,s = 7,phi = fit_7$phi, vara = fit_7$avar)

# Set the number of realizations
sims <- 1

# Generate and plot the first realization to establish the plot
S10AR3gen <- gen.arima.wge(319, s = 7, phi = fit_7$phi, vara = fit_7$avar)
plot(S10AR3gen, type = "l", lwd = 2)

# Generate and plot additional realizations in different colors
for(i in 1:sims) 
{

   S10AR3gen_new <- gen.arima.wge(319, s = 7, phi = fit_7$phi, vara = fit_7$avar, plot = "FALSE")
   lines(S10AR3gen_new,col = "red", lwd = 2)
}

plotts.sample.wge(S10AR3gen)

```

###  Zoom in the last 182 values to see how the generated realizations fit.

```{r}

# Assuming BikeSharing$total_cust is your time series object
# Select the last 182 values <=> about 26 weeks = 182 days
last_26weeks_values <- tail(BikeSharing$total_cust, 182)

# Now you can use last_60_values in your analysis or plotting
# For example, if you want to plot these values:
plot.ts(last_26weeks_values)

# Set the number of realizations
sims <- 1
# Generate and plot the first realization to establish the plot
S10AR3gen <- gen.arima.wge(319, s = 7, phi = fit_7$phi, vara = fit_7$avar)
plot(tail(S10AR3gen, 182), type = "l", lwd = 3) # Plot only the last 60 values

# Generate and plot additional realizations in red
for(i in 1:sims) {
   S10AR3gen_new <- gen.arima.wge(319, s = 7, phi = fit_7$phi, vara = fit_7$avar, plot = "FALSE")
   lines(tail(S10AR3gen_new, 182), col = "red", lwd = 2) # Plot only the last 60 values
}

```

# VAR Model

```{r}
no_missing_values = apply(BikeSharing, 2, function(x) all(complete.cases(x)))
columns_no_missing = names(BikeSharing)[no_missing_values]

# All columns
colnames(BikeSharing)
# Columns with no missing values
print(columns_no_missing)
# date, total_cust, temp_min, temp_max, temp_observ, precip, wind
```

```{r}
data <- data.frame(x = is.na(BikeSharing$wt_rain))

gg <- ggplot(data, aes(x = 1:nrow(data), y = x)) +
  geom_point(size = 1, height = .5, width = .5, color = "red") +
  xlab("days by row number") +
 ylab("wt_rain Values Present or Missing") +
 labs(title = "Scatter Plot of Missing wt_rain Values", size = 10) +
 scale_y_discrete(labels = c("Present", "Missing")) +
 scale_x_continuous(breaks = seq(0, nrow(data), by = 100)) +
 theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
 theme_classic()
gg
```

```{r}
# Select columns with no missing values excluding 'date' and 'precip_L1'
variables = c("total_cust", "temp_min", "temp_max", "temp_observ", "precip", "wind", "holiday")
BikeSharingSelected = BikeSharing[variables]

# Leave out the last 182 values for the ASE calculation
BikeSharingSmall = head(BikeSharingSelected, -182)

# Prepare the data for the VAR model
BikeSharingDF = data.frame(BikeSharingSmall[,-1]) # Exclude 'total_cust' for independent variables
TotalCust = BikeSharingSmall$total_cust
```

```{r}
# Double check we have all the dependent variables correctly saved to the dataframe
head(BikeSharingDF)
```

```{r}
VAR_Select_Bikeshare = VARselect(cbind(TotalCust, BikeSharingDF),lag.max = 10, type = "both")
VAR_Select_Bikeshare
```

```{r}
# Create the VAR model
VAR_BikeSharing = VAR(cbind(TotalCust, BikeSharingDF), lag.max = 10, type = "both", p = 9)

# Prediction
pred_VAR_7 = predict(VAR_BikeSharing, n.ahead = 7)
pred_VAR_182 = predict(VAR_BikeSharing, n.ahead = 182)
```

```{r}
# Plot the actual vs predicted 'total_cust'
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted 1 Week Forecast VAR", xlab = "Time", ylab = "Total Customers", xlim = c(2800, nrow(BikeSharing)))
lines((nrow(BikeSharing) - 6):nrow(BikeSharing), pred_VAR_7$fcst$TotalCust[,1], col = "red")
```

```{r}
# Plot the actual vs predicted 'total_cust' zoomed in on the last 1/6th of the timeline
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted 26 Week Forecast VAR", xlab = "Time", ylab = "Total Customers", xlim = c(2500, nrow(BikeSharing)))
lines((nrow(BikeSharing) - 181):nrow(BikeSharing), pred_VAR_182$fcst$TotalCust[,1], col = "red")
```

```{r}
# Calculate the Average Squared Error (ASE) for the prediction
ASE_VAR_7 = mean((tail(BikeSharing$total_cust, 7) - pred_VAR_7$fcst$TotalCust[,1])^2)
print(ASE_VAR_7)

ASE_VAR_182 = mean((tail(BikeSharing$total_cust, 182) - pred_VAR_182$fcst$TotalCust[,1])^2)
print(ASE_VAR_182)
```

# MLP Model
```{r}
# Exclude the first 2,300 values from BikeSharingSmall and BikeSharingDF
BikeSharingSmallAdjusted = BikeSharingSmall[-(1:2300), ]
BikeSharingDFAdjusted = BikeSharingDF[-(1:2300), ]
```

```{r}
fit_mlp = mlp(ts(BikeSharingSmallAdjusted$total_cust), reps = 7, comb = "median", xreg = BikeSharingDFAdjusted)
plot(fit_mlp)
```

```{r}
fore_mlp_182 = forecast(fit_mlp, h = 182, xreg = BikeSharingSelected[-(1:2300), -1])
fore_mlp_7 = forecast(fit_mlp, h = 7, xreg = BikeSharingSelected[-(1:2300), -1])
```

```{r}
plot(fore_mlp_182)
```

```{r}
# Plot the actual 'total_cust' values
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted 26 Week Forecast MLP", xlab = "Time", ylab = "Total Customers", xlim = c(2500, nrow(BikeSharing)))

# Add the MLP forecasted 'total_cust' values
lines((nrow(BikeSharing) - 181):nrow(BikeSharing), fore_mlp_182$mean, col = "red")
```

```{r}
# Plot the actual 'total_cust' values
plot(BikeSharing$total_cust, type = "l", main = "Total Customers: Actual vs Predicted 1 Week Forecast MLP", xlab = "Time", ylab = "Total Customers", xlim = c(2800, nrow(BikeSharing)))

# Add the MLP forecasted 'total_cust' values
lines((nrow(BikeSharing) - 6):nrow(BikeSharing), fore_mlp_7$mean, col = "red")
```

```{r}
ASE_mlp_182 = mean((tail(BikeSharing$total_cust, 182) - fore_mlp_182$mean)^2)
print(ASE_mlp_182)

ASE_mlp_7 = mean((tail(BikeSharing$total_cust, 7) - fore_mlp_7$mean)^2)
print(ASE_mlp_7)
```

# Ensemble

```{r}
fit_ar = est.ar.wge(BikeSharing$total_cust, p = 8)
preds_ar_182 = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_ar$phi, n.ahead = 182,
    lastn = TRUE, limits = FALSE)

#Ensemble
ensemble_182 = (pred_VAR_182$fcst$TotalCust[, 1] + fore_mlp_182$mean + preds_ar_182$f) / 3

#Plot
plot(BikeSharing$total_cust, type = "l", xlim = c(2500, nrow(BikeSharing)), ylab = "Total Customers", main = "Total Customers: Actual vs Predicted 26 Week Customer Forecast Ensemble")
lines((nrow(BikeSharing) - 181):nrow(BikeSharing), ensemble_182, type = "l", col = "red")

ASE_ensemble_182 = mean((tail(BikeSharing$total_cust, 182) - ensemble_182)^2)
ASE_ensemble_182
```

```{r}
preds_ar_7 = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_ar$phi, n.ahead = 7,
    lastn = TRUE, limits = FALSE)

#Ensemble
ensemble_7 = (pred_VAR_7$fcst$TotalCust[, 1] + fore_mlp_7$mean + preds_ar_7$f) / 3

#Plot
plot(BikeSharing$total_cust, type = "l", xlim = c(2800, nrow(BikeSharing)), ylab = "Total Customers", main = "Total Customers: Actual vs Predicted 1 Week Customer Forecast Ensemble")
lines((nrow(BikeSharing) - 6):nrow(BikeSharing), ensemble_7, type = "l", col = "red")

ASE_ensemble_7 = mean((tail(BikeSharing$total_cust, 7) - ensemble_7)^2)
ASE_ensemble_7
```