---
title: "Time Series Project"
author: "Derek Rogers and Sakava Kiv"
date: "2024-04-06"
output: html_document
---


```{r}
library(tswge)
library(PolynomF)
library(tidyverse)
library(GGally)
library(astsa)
library(vars)
library(nnfor)
```

```{r}
# Set the working directory to the specified path
#setwd("C:/SMU MS Data Science/TimeSeries/FinalProject/BikeSharingTimeSeriesAnalysis")

# Check to confirm that the working directory has been set correctly
#getwd()
```

```{r}
# The chosen data set is the bike_sharing_dataset.csv from kaggle: https://www.kaggle.com/datasets/juliajemals/bike-sharing-washington-dc
# The chosen response variable is total_cust
# Being able to model and predict total_cust can help entities directly involved or adjacent to bike sharing better understand and predict the varying demand to avoid oversupply or shortages. Forecasting future demand can help plan bike maintance as heavily used bikes break down more often. This information can help inform business decisions, setting prices, and creating efective advertisements. 

BikeSharing = read.csv("bike_sharing_dataset.csv", header = TRUE)
BikeSharing_train = data.frame(TotalCust = ts(BikeSharing$total_cust))

head(BikeSharing)
nrow(BikeSharing)

# There are 4 consecutive missing values
which(is.na(BikeSharing$total_cust))
```

```{r}
# Impute the 4 consecutive missing values with the average of the value before and after the missing values

# Calculate the average of the values at positions 1848 and 1853
average_value = mean(c(BikeSharing$total_cust[1848], BikeSharing$total_cust[1853]), na.rm = TRUE)
average_value

length(BikeSharing$total_cust)

# Assign the average value to the missing positions 1849, 1850, 1851, and 1852
BikeSharing$total_cust[1849:1852] = average_value

# Optionally, you can verify the imputation by checking the values at these positions
BikeSharing$total_cust[1848:1853]

# Plot around where the imputed values are
plot(BikeSharing$total_cust[1838:1863], type = "l")
```
Plotting the full dataset
The realization plot shows strong periodic trends, wandering behavior, and variance that changes over time. We see slowly dampening autocorrelation with weak and an irregular length periodic behavior. There is a peak at zero in the parzen frequency window (spectral density plot),This could suggest a yearly seasonal trend since we have 1/365 = 0.0027.The parzen window also shows a peak around 0.15, which is suggest a period of 1/7 = 0.1428, This number is close to the .15 estimate we see in the second highest peak from the spectral density plot. Because this data is collected daily, there evidence to suggest a weekly trend. We can hypothesize total_cust could have a greater total_cust on the weekends when all other variables are held constant. Our intuition also suggest ridership is daily, since people will bike continue to bike the next day if that is the main mode of transportation or for exercise purposes.Another thing to note we see evidence of non stationary data as it seems the trend for total_cust increases over time cyclically as and does matter where we are in time series time point.Also note since we see slowly damping sample auto correlations, this suggest the need to difference the data. 

```{r}
#plot sample of the data 
plotts.sample.wge(BikeSharing$total_cust)

#step 1 visual test
plot.ts(BikeSharing$total_cust)

head(BikeSharing)

# Assuming BikeSharing$total_cust is a numeric vector with the total customers per time point
# Create a sequence of dates starting from 2011-01-01
dates <- seq(as.Date("2011-01-01"), by="day", length.out=length(BikeSharing$total_cust))

# Combine the dates and total customers into a data frame
BikeSharing_df <- data.frame(Date=dates, Total_Customers=BikeSharing$total_cust)

# Convert the data frame to a ts object
BikeSharing_ts <- ts(BikeSharing_df$Total_Customers, start=c(2011, 1), frequency=365)


# Now plot the time series to see the 5 year period trends
plot.ts(BikeSharing_ts)
```

We will try to experiment to estimate for candidate AR or ARMA model based on AIC and BIC values here. We can observe that AN AR(8) model was picked by both AIC AND BIC and also by the PACF function we saw evidence of an AR(8) as well.

```{r}
# Top AIC was p = 9, q = 4
aic5.wge(BikeSharing$total_cust, p = 0:12, q = 0:4)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:0)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:0)


# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:12, q = 0:4,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:2,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:9, q = 0:0,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:2,type = "bic")
# Top BIC was p = 2, q = 1
aic5.wge(BikeSharing$total_cust, p = 0:2, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(BikeSharing$total_cust, p = 0:8, q = 0:0,type = "bic")


pacf(BikeSharing$total_cust) # suggest an AR(8) model

```

Observe that from the code above by running the estimate for p and q for both AIC and BIC for ranges of P and Q of my choosing that make sense, I will probably go with an AR(8) since this seems to strike a balance between the AIC and BIC values. 

And checking with PACF function we see evidence of an AR(8) model. 

Next We will see what happens if I apply a transformation for S=7 and S=365 to see if these are useful.  

Note to self: 
It is not a good idea to just look at the factor tables. Look at the data, look at the realizations, look at the ACFs, and look at the factor tables, and together you can gather a lot of information to find the most useful model. So before and after differencing. So you should check all these out before and after differences.


```{r}

# transform data
# Difference the data to until we get a stationary model to check for (1-B)^d
# we could use the Dickey-Fuller Test he to see if model as evidence for a unit root, but differencing here I do not see visual evidence for a (1-B) term.
b1=artrans.wge(BikeSharing$total_cust,phi.tr=1)
b2=artrans.wge(b1,phi.tr=1)
b3=artrans.wge(b2,phi.tr=1)
#difference the data for (1-B^7) fit
b7=artrans.wge(b1,phi.tr = c(rep(0,6),1))

plotts.sample.wge(b1, arlimits = TRUE)
plotts.sample.wge(b7, arlimits = TRUE)
#from the above there does not seem to be evidence for a (1-B)^d So we will not include this in the model. 
#The Autocorrelation Function (ACF) plot does not suggest white noise. In a white noise series, we would expect most of the spikes in the ACF plot to fall within the confidence interval (usually represented by blue shaded areas or dashed lines), indicating that there is no significant correlation at different lags.

#However, the acf indicates that there are significant correlations at various lags. This suggests that the time series data exhibits some patterns or relationships between its values at different times, which is not characteristic of white noise.

# transform data
# Difference the data to until we get a stationary model to check for (1-B^S) i.e (1-B^7) and (1-B^365)

#difference the data for (1-B^7) fit
b7=artrans.wge(BikeSharing$total_cust,phi.tr = c(rep(0,6),1))

#difference the data furthur with (1-B^365) fit
b7_365=artrans.wge(b7,phi.tr = c(rep(0,364),1)) #after doing this the data appears to be more like white noise indicating a stationary model so adding a (1-B^7) and (1-B^365) to our model would be useful.
b7_365_1=artrans.wge(b7_365,phi.tr=1)
# estimate parameters of stationary part
est.7.365=est.ar.wge(b7_365,p=8)
#est.7.365.1=est.ar.wge(b7_365_1,p=8)
#est.7.365
est.7=est.ar.wge(b7,p=8)
est.7.365$res

```

```{r}

# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:12, q = 0)
# Top AIC was p = 7, q = 2
aic5.wge(b7, p = 0:9, q = 0:2)
# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:9, q = 0:1)
# Top AIC was p = 9, q = 0
aic5.wge(b7, p = 0:9, q = 0:0)
# Top AIC was p = 7, q = 2
aic5.wge(b7, p = 0:8, q = 0:2)
# Top AIC was p = 8, q = 1
aic5.wge(b7, p = 0:8, q = 0:1)
# Top AIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:0)


# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:12, q = 0,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:2,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:9, q = 0:0,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:2,type = "bic")
# Top BIC was p = 1, q = 0
aic5.wge(b7, p = 0:2, q = 0:1,type = "bic")
# Top BIC was p = 8, q = 0
aic5.wge(b7, p = 0:8, q = 0:0,type = "bic")


pacf(BikeSharing$total_cust) # suggest an AR(8) model

```


```{r}

#AIC and BIC Selects an AR(8), and the final model is:
# phi_subs8(B)(1-B^7)X_t = at,
# Where phisubs8(B) = 1-0.4043 - 0.0180B^2 - 0.0225B^3- 0.0428B^4 + 0.0060B^5 - 0.0304B^6 + 0.5069B^7 - 0.2151B^8
# and Sigmahat_a^2 = 4374309 
est.7=est.ar.wge(b7,p=8, method = 'burg')

#step 1 visual test => data seems to be non stationary
plot.ts(est.7$res)


#from this sample plot we can see in the ACF evidence for s=7, since we see peaks at the lags 7,14, and around 21
plotts.sample.wge(est.7$res, arlimits = TRUE)

# get the phis 
est.7$phi

#run ljung test for white noise
ljung.wge(est.7$res,p=8)

ljung.wge(est.7$res,K=48,p=8)


#Forecasting with SARIMA model here such that ARIMA(p,d,q) with s=7
f = fore.arima.wge(BikeSharing$total_cust, s=7, phi = est.7$phi, n.ahead = 182, limits =  F, lastn = T, plot=TRUE)

f

ASE = mean((BikeSharing$total_cust[(length(BikeSharing$total_cust)-181):length(BikeSharing$total_cust)] - f$f)^2)

ASE

#USE box and Jenkins Method to find what type of AR model this is with the PACF function.
pacf(BikeSharing$total_cust) # suggest an AR(8) model

#compare overfit estimated table with the season factor table of S =7 => observe that when comparing the tables we see some evidence of an S=7, although the estimates are not that strongly matching the (1-B^7) factor tables
esttable=est.ar.wge(BikeSharing$total_cust,p = 12,method='burg') #overfit for seasonality s= 7
# estimate parameters of stationary part
b7table = factor.wge(c(rep(0,6),1)) #1-B^7 table
#factor.wge(c(rep(0,9),1)) #1-B^10


#Compare Spectral Densities
sims = 30
SpecDen = parzen.wge(BikeSharing$total_cust, plot = "FALSE")
plot(SpecDen$freq,SpecDen$pzgram, type = "l", lwd = 6)

for( i in 1: sims)
{
   SpecDen2 = parzen.wge(gen.arima.wge(319, s=7, phi = est.7$phi, plot ="FALSE"), plot ="FALSE")
   lines(SpecDen2$freq,SpecDen2$pzgram, lwd = 2, col = "red")
}


#Compare ACFs
sims = 30
ACF = acf(BikeSharing$total_cust, plot = "FALSE")
plot(ACF$lag ,ACF$acf , type = "l", lwd = 6)

for( i in 1: sims)
{
   ACF2 = acf(gen.arima.wge(319, s = 7, phi = est.7$phi, plot ="FALSE"), plot ="FALSE")
   lines(ACF2$lag ,ACF2$acf, lwd = 2, col = "red")
}

#Compare Generated Realizations 
S10AR3gen = gen.arima.wge(319,s = 7,phi = est.7$phi, vara = est.7$avar)


plotts.sample.wge(S10AR3gen)



# Set the number of realizations
sims <- 1

# Generate and plot the first realization to establish the plot
S10AR3gen <- gen.arima.wge(319, s = 7, phi = est.7$phi, vara = est.7$avar)
plot(S10AR3gen, type = "l", lwd = 2)

# Generate and plot additional realizations in different colors
for(i in 1:sims) 
{

   S10AR3gen_new <- gen.arima.wge(319, s = 7, phi = est.7$phi, vara = est.7$avar, plot = "FALSE")
   lines(S10AR3gen_new,col = "red", lwd = 2)
}



# Assuming BikeSharing$total_cust is your time series object
# Select the last 182 values <=> about 6 months = 182 days
last_6mon_values <- tail(BikeSharing$total_cust, 182)

# Now you can use last_60_values in your analysis or plotting
# For example, if you want to plot these values:
plot.ts(last_6mon_values)

# Set the number of realizations
sims <- 1
# Generate and plot the first realization to establish the plot
S10AR3gen <- gen.arima.wge(319, s = 7, phi = est.7$phi, vara = est.7$avar)
plot(tail(S10AR3gen, 182), type = "l", lwd = 3) # Plot only the last 60 values

# Generate and plot additional realizations in different colors
for(i in 1:sims) {
   S10AR3gen_new <- gen.arima.wge(319, s = 7, phi = est.7$phi, vara = est.7$avar, plot = "FALSE")
   lines(tail(S10AR3gen_new, 182), col = "red", lwd = 2) # Plot only the last 60 values
}





```

```{r}
# Find ASE for a 6 month forecast
fit_ar = est.ar.wge(BikeSharing$total_cust, p = 8)
preds_arma = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_ar$phi, n.ahead = 182,
    lastn = TRUE, limits = FALSE)

ASE = mean((tail(BikeSharing$total_cust, 182) - preds_arma$f)^2)
ASE


# Sakava Alternative code with S = 7
f6m = fore.arima.wge(BikeSharing$total_cust, s=7, phi = est.7$phi, n.ahead = 182, limits =  F, lastn = T, plot=TRUE)


ASE = mean((BikeSharing$total_cust[(length(BikeSharing$total_cust)-181):length(BikeSharing$total_cust)] - f6m$f)^2)

ASE

#more concise way to calculate ASE
ASE = mean((tail(BikeSharing$total_cust, 182) - f6m$f)^2)

ASE
```


```{r}
# Find ASE for a 1 week forecast
fit_arma = est.arma.wge(BikeSharing$total_cust, p = 9, q = 4)
preds_arma = fore.arma.wge(BikeSharing$total_cust,
    phi = fit_arma$phi, theta = fit_arma$theta, n.ahead = 7,
    lastn = TRUE, limits = FALSE)

ASE = mean((tail(BikeSharing$total_cust, 7) - preds_arma$f)^2)
ASE


# Sakava Alternative code with S = 7
f1w = fore.arima.wge(BikeSharing$total_cust, s=7, phi = est.7$phi, n.ahead = 7, limits =  F, lastn = T, plot=TRUE)

ASE = mean((BikeSharing$total_cust[(length(BikeSharing$total_cust)-7):length(BikeSharing$total_cust)] - f1w$f)^2)

ASE


```

```{r}

```

```{r}

```

```{r}

```

```{r}

```



